---
title: "Topic Modelling of New York Times Article Headlines During the Spanish Flu & COVID-19 Pandemic"
subtitle: "Term paper SURV727"
author: "Anthony Garove and Ujjayini Das"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    df_print: kable
references:
- id: Wickham2014
  title: Tidy Data
  author:
  - family: Wickham
    given: Hadley
  container-title: Journal of Statistical Software
  volume: 59
  issue: 10
  page: 1-23
  type: article-journal
  issued:
    year: 2014
- id: Baumer2017
  title: Modern Data Science with R
  author:
  - family: Baumer
    given: Benjamin S.
  - family: Kaplan
    given: Daniel T.
  - family: Horton
    given: Nicholas J.
  type: book
  publisher: Chapman \& Hall/CRC Press.
  issued:
    year: 2017
---

```{r, include = FALSE}
library(knitr)
library(tidyverse)
```

## Introduction

This section outlines the research idea. We can also cite related work here [@Wickham2014; @Baumer2017].

Note that compiled term paper (the PDF) is supposed to be more text-centered than the RMarkdown documents we used in class, i.e. the text sections are more detailed and big or redundant code chunks can be hidden.

  For our final term paper, Ujjayini and I were generally interested in making historical comparisons across different pandemics. Our overarching purpose was to obtain insights about the similarities in major societal issues caused by pandemics, as well as to examine time-era specific differences across pandemics from different points in history.
  
  We specifically chose to compare the Spanish Flu Pandemic of 1918 to the modern COVID-19 pandemic. We were curious to see how two pandemics from different centuries differed in terms of the areas of life they affected across the globe. We also considered that the type of disease in each pandemic may cause differences in how they affected the world (i.e., an influenza versus a coronavirus pandemic). We thought that searching the historical archives of printed media may provide insights to the research question at hand.
  
  We operationalized this research question by asking, "using Latent Dirichlet Allocation, are there differences in the topics that emerge from published New York Times article headlines during the first respective month of the Spanish Flu and COVID-19 pandemic?"

## Data

This section describes the data sources and the data gathering process.

 We used the New York Times Archive API to access N=12,028 article headlines published during the first month of each pandemic. To the best of our knowledge, the Spanish Flu was officially declared a pandemic on 4th March, 1918. The COVID-19 public health crisis was officially declared a pandemic on 11th March, 2020.

 We used the specific endpoints, month and year, while requesting for published article headlines from the New York Times Archive API with an authorised key. The following code was used to retrieve the data:

```{r}
# Building URLs
base_url1 <- 'https://api.nytimes.com/svc/archive/v1/1918/3.json?api-key=SVyAXhJrhFVCMF4tvGU2GZY3jm1greFU'
request1 <- GET(base_url1)
base_url2 <- 'https://api.nytimes.com/svc/archive/v1/2020/3.json?api-key=SVyAXhJrhFVCMF4tvGU2GZY3jm1greFU'
request2 <- GET(base_url2)

# Getting the Contents as Data Frames
response1 <- content(request1, as = "text", encoding = "UTF-8")
spanishfludata <- fromJSON(response1, flatten = TRUE) %>% data.frame()
response2 <- content(request2, as = "text", encoding = "UTF-8")
coviddata <- fromJSON(response2, flatten = TRUE) %>% data.frame()

# Extract the Subsetted Data 
headline_spanish <- spanishfludata[,19]
headline_covid <- coviddata[,20]
```

  We first formatted the JSON objects into two data frames --- one for the Spanish Flu and another for the COVID-19. After that, we extracted the headlines from both of the data frames and used those extracted objects for the remaining analysis.

## Results

### Pre Processing 

  Next, we pre-processed the headline texts using standard text mining procedures in order to prepare the data for topic modelling. These processes included tokenization, lemmatization and removing stopwords.Tokenization is a method used in natural language processing that separates text data into units called "tokens". Tokenization can be done to break up data in to words, characters, and subwords. We used n=1 gram tokenization, which separates our headline data into one-word texts. 
  
  Lemmatization is a process that normalizes text such that words are substituted with their base root. This step is used to categorize words into groups of root forms with similar meanings. Stopwords are words that are frequently used in a given language, and these words are removed prior to topic modelling. This step is necessary because stopwords contribute noise when performing Latent Dirichlet Allocation, and removing stopwards allows researchers to interpret the topics extracted from text with more coherence. For both lemmatization and removal of stopwords, we used the standard English dictionary lemmas and stopwords.

  After performing these pre-processing techniques, two corpora of texts were produced; one for Spanish Flu headlines and another for COVID-19 headlines. Based on the two corpora, we then created two document term matrices---where an individual row represents a single document, a column represents a single term, and each value in a matrix contains the frequency of that term in the document. Document-term-matrices are used as inputs for topic modelling with the `topicmodels` package. The code below was used for tokenization, lemmatization, removal of stopwards, and creation of document-term-matrices:

```{r}
# Tokenization
headline_spanish_gsub <- gsub("'","",headline_spanish)
headline_covid_gsub <- gsub("’","",headline_covid)
token_spanish <- tokens(headline_spanish_gsub,
                        remove_punct = TRUE,
                        remove_numbers = TRUE,
                        remove_symbols = TRUE) %>%
  tokens_tolower()
token_covid <- tokens(headline_covid_gsub,
                        remove_punct = TRUE,
                        remove_numbers = TRUE,
                        remove_symbols = TRUE) %>%
  tokens_tolower()

# Lemmatization
library(haven)
lemmaData <- read_sav("./lemma_spss.sav")

corpus_spanish <-  tokens_replace(token_spanish, # "Substitute token types based on vectorized one-to-one matching"
                                    lemmaData$inflected_form, 
                                    lemmaData$lemma,
                                    valuetype = "fixed") 
corpus_covid <-  tokens_replace(token_covid, # "Substitute token types based on vectorized one-to-one matching"
                                    lemmaData$inflected_form, 
                                    lemmaData$lemma,
                                    valuetype = "fixed") 

# Removing Stopwords
corpus_spanish <- corpus_spanish %>%
                             tokens_remove(stopwords("english")) %>%
                             tokens_ngrams(1)
corpus_covid <- corpus_covid %>%
                             tokens_remove(stopwords("english")) %>%
                             tokens_ngrams(1)

#Create document term matrices
DTM_spanish <- dfm(corpus_spanish)
minimumFrequency <- 10
DTM_spanish <- dfm_trim(DTM_spanish, 
                min_docfreq = minimumFrequency,
                max_docfreq = 1000000)
DTM_spanish  <- dfm_select(DTM_spanish, 
                   pattern = "[a-z]", 
                   valuetype = "regex", 
                   selection = 'keep')
colnames(DTM_spanish) <- stringi::stri_replace_all_regex(colnames(DTM_spanish), 
                                                 "[^_a-z]","")
DTM_spanish <- dfm_compress(DTM_spanish, "features")
sel_idx_spanish <- rowSums(DTM_spanish) > 0
DTM_spanish <- DTM_spanish[sel_idx_spanish, ]
corpus_spanish <- corpus_spanish[sel_idx_spanish, ]
DTM_covid <- dfm(corpus_covid)
minimumFrequency <- 10
DTM_covid <- dfm_trim(DTM_covid, 
                min_docfreq = minimumFrequency,
                max_docfreq = 1000000)
DTM_covid  <- dfm_select(DTM_covid, 
                   pattern = "[a-z]", 
                   valuetype = "regex", 
                   selection = 'keep')
colnames(DTM_covid) <- stringi::stri_replace_all_regex(colnames(DTM_covid), 
                                                 "[^_a-z]","")
DTM_covid <- dfm_compress(DTM_covid, "features")
sel_idx_covid <- rowSums(DTM_covid) > 0
DTM_covid <- DTM_covid[sel_idx_covid, ]
corpus_covid <- corpus_covid[sel_idx_covid, ]
```

### Topic Modeling Using Latent Dirichlet Allocation

  We used Latent Dirichlet Allocation (LDA) to examine topics from our two corpora of headlines. LDA is a frequently used algorithm for fitting a topic model. LDA is based on an assumption that documents used in your analysis are comprised of a mixture of topics, and that every topic is an amalgamation of words. LDA simultaneously finds the words associated with each topic while extracting the topics contained in each document.
  
  We specified our LDA model to extract 

```{r}
K_spanish <- 15
set.seed(1234)
topicModel_spanish2 <- LDA(DTM_spanish, 
                   K_spanish, 
                   method="Gibbs", 
                   control=list(iter = 500, 
                                verbose = 25, 
                                alpha = 0.2))
tmResult_spanish2 <- posterior(topicModel_spanish2)
theta_spanish2 <- tmResult_spanish2$topics
beta_spanish2 <- tmResult_spanish2$terms
topicProportions_spanish2 <- colSums(theta_spanish2) / nrow(DTM_spanish)  # average probability over all paragraphs
names(topicProportions_spanish2) <- topicNames_spanish     # Topic Names 
sort(topicProportions_spanish2, decreasing = TRUE) # sort
topicNames_spanish2 <- apply(terms(topicModel_spanish2, 5), 2, paste, collapse = " ")  # top five terms per topic 


```

```{r}
# What happens here depends on the specific project
```

```{r}
# What happens here depends on the specific project
```

## Discussion

This section summarizes the results and may briefly outline advantages and limitations of the work presented.

## References
